---
title: "Spotify 2023 Data Analysis"
author: "Shahrukh Rahman"
date: "`r Sys.Date()`"
output: html_document
---

# **Introduction**
This report analyzes the **Spotify 2023 dataset**, performing various data manipulation, visualization, and statistical computations in R.

```{r setup, include=FALSE}
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# **1. Load and View Dataset**

```{r}
spotify_2023_Copy <- read_excel("spotify-2023 - Copy.xls")
View(spotify_2023_Copy)
```

# **2. List the variables in your dataset **
```{r}
str(spotify_2023_Copy)  # Check structure of dataset
colnames(spotify_2023_Copy)  # Display column names
```
# **3.Top 15 rows of your dataset **
```{r}
head(spotify_2023_Copy, 15)  # Display first 15 rows
```

# **4. User defined function **
## **Categorize Streams**
```{r}
categorize_streams <- function(streams) {
  if (streams >= 1000000) {
    return("High Stream")
  } else {
    return("Low Stream")
  }
}
spotify_2023_Copy$Stream_Category <- sapply(spotify_2023_Copy$streams, categorize_streams)
head(spotify_2023_Copy)
```

## **5. Filtering Recent Songs**
```{r}
recent_songs <- spotify_2023_Copy %>% filter(released_year > 2020)
head(recent_songs)
```

## **6. Reshape Data**
```{r}
spotify_2023_Copy <- spotify_2023_Copy %>%
  rename(artist_name = `artist(s)_name`) %>%  # Rename artist column
  select(track_name, artist_name, streams, bpm, `danceability_%`, `energy_%`) %>%  # Select columns
  pivot_longer(
    cols = c(bpm, `danceability_%`, `energy_%`),  # Use backticks for special names
    names_to = "Feature",
    values_to = "Value"
  )

head(spotify_2023_Copy)
```

# **7. Data Cleaning**
## **Remove Missing Values**
```{r}
cleaned_df <- spotify_2023_Copy %>% drop_na()
sum(is.na(cleaned_df))  # Check if missing values are removed
```

## **8. Remove Duplicated Rows**
```{r}
duplicated_rows <- spotify_2023_Copy %>% filter(duplicated(spotify_2023_Copy))
unique_df <- spotify_2023_Copy %>% distinct()
```

## **9. Reorder Rows in Descending Order**
```{r}
sorted_df <- spotify_2023_Copy %>% arrange(desc(streams))
head(sorted_df)
```

# **10. Data Transformation**
## **Renaming  Column Name**
```{r}
spotify_2023_Copy <- spotify_2023_Copy %>% rename(artis_name = `artist_name`)
colnames(spotify_2023_Copy)
```

## **11. Creating a New Variable Using Mathematical Function**
```{r}
spotify_2023_Copy <- spotify_2023_Copy %>%
  mutate(Value = ifelse(Feature == "bpm", as.numeric(Value) * 2, Value))



```

# **12. Create a Training Set**
```{r}
set.seed(123)
train_index <- sample(1:nrow(spotify_2023_Copy), 0.8 * nrow(spotify_2023_Copy))
train_set <- spotify_2023_Copy[train_index, ]
test_set <- spotify_2023_Copy[-train_index, ]
```

```{r}
nrow(train_set)  # Check training set size
nrow(test_set)   # Check test set size
```

# **13. Summary Statistics**
```{r}
summary(spotify_2023_Copy)
```

## **14. Statistical Functions**
```{r}
spotify_2023_Copy <- spotify_2023_Copy %>%
  mutate(streams = gsub("[^0-9]", "", streams),  # Remove non-numeric characters
         streams = as.numeric(streams))         # Convert to numeric

mean(spotify_2023_Copy$streams, na.rm = TRUE)
median(spotify_2023_Copy$streams, na.rm = TRUE)

```

```{r}
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
get_mode(spotify_2023_Copy$streams)
```

```{r}
range(spotify_2023_Copy$streams, na.rm = TRUE)
```

# **15. Data Visualization**
## **Scatter Plot (BPM vs Energy)**
```{r}
# Summarize duplicates by averaging the Value column
spotify_summarized <- spotify_2023_Copy %>%
  group_by(track_name, artis_name, streams, Feature) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop")

# Now pivot the data
spotify_wide <- spotify_summarized %>%
  pivot_wider(names_from = Feature, values_from = Value)

# Convert necessary columns to numeric
spotify_wide <- spotify_wide %>%
  mutate(bpm = as.numeric(bpm), `energy_%` = as.numeric(`energy_%`))

# Check the result
head(spotify_wide)


ggplot(spotify_wide, aes(x = bpm, y = `energy_%`)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of BPM vs Energy",
       x = "Beats Per Minute (BPM)",
       y = "Energy Percentage") +
  theme_minimal()

```

## **16. Bar Plot (Released Year vs Spotify Playlists)**
```{r}
spotify_summary <- spotify_2023_Copy %>%
  group_by(Feature) %>%  # Grouping by the 'Feature' column
  summarise(total_streams = sum(streams, na.rm = TRUE), .groups = "drop")

ggplot(spotify_summary, aes(x = Feature, y = total_streams)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Bar Plot of Feature vs Total Streams",
       x = "Feature",
       y = "Total Streams") +
  theme_minimal()



```

# **17. Correlation Analysis**
```{r}
library(dplyr)
library(tidyr)

# Summarize duplicates by calculating the mean for each combination
spotify_filtered <- spotify_2023_Copy %>%
  filter(Feature %in% c("bpm", "energy_%")) %>%
  group_by(track_name, artis_name, streams, Feature) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Feature, values_from = Value)

# Check the data to ensure bpm and energy_% are in separate columns
head(spotify_filtered)

spotify_filtered$bpm <- as.numeric(spotify_filtered$bpm)
spotify_filtered$`energy_%` <- as.numeric(spotify_filtered$`energy_%`)

# Calculate the correlation
correlation_value <- cor(spotify_filtered$bpm, spotify_filtered$`energy_%`, use = "complete.obs", method = "pearson")
print(correlation_value)



```

# **9. Save and Load Session**
```{r}
# Saving the session in the specified directory
save.image("C:/Users/anjal/Downloads/Spotify v2/Spotify v2/Spotify_Session.RData")
# Loading the session from the specified directory
load("C:/Users/anjal/Downloads/Spotify v2/Spotify v2/Spotify_Session.RData")


